RL-Musician: A Tool for Music Composition with Reinforcement Learning

### 2019-11-03 docs/paper/paper.tex
docs: Fix minor issues

\date{Ocotber 26, 2019}

@@

\keywords{algorithmc composition \and music generation \and reinforcement learning}

@@

Currently, there are no perfect tools for algorithmic composition. Plenty of various approaches for composing music automatically exist, but none of them produce well-structured and novel pieces that can not be distinguished from works of a talented human composer. Thus, there is an open research problem. Several recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating, payne2019musenet} and reinforcement learning \cite{smith2012reinforcement} are applicable. Moreover, some researchers combine them \cite{jaques2016generating, kotecha2018bach, kumar2019polyphonic}.

@@

Agent takes action by drawing it from probability distribution returned by its actor model. An actor model can be every function that takes observation as input and returns probability distribution over cells from the current time step. Here, shallow neural network is used as actor model. Its weights can be viewed as something that defines transition matrix for a Markov chain from which sequence of piano roll's states is generated.

@@

\begin{itemize}
	\item \textbf{Consonances.} Speaking formally, total reward for consonances is:
	$$\sum_{t = 1}^m \sum_{i, j = 1}^n [i > j]R_{it} R_{jt} c((i - j) \bmod 12),$$
	where square brackets are Iverson brackets and $c$ is a function that maps interval in semitones into measure of its consonance, $c: \{0, \dots, 11\} \to \mathbb{R}$.
	The interpretation is as follows. Since $R_{it} \in \{0, 1\}$, only pairs of sounding cells from the same column affect the total sum. Value of each non-zero term depends on interval between the two corresponding pitches. The more consonant the interval is, the higher the value should be. Function $c$ must be chosen in a way providing that. However, there is an exception -- here, $c(0) = 0$ although intervals of whole number of octaves are perfectly consonant. This is done in order to prevent doubling the same melody in different octaves. 
	\item \textbf{Absence of notes not from the specified scale.} Mode (major or minor) and position of a row with the tonic are set by user before training. The same negative penalty is assigned to each sounding cell of a pitch that does not belong to the specified scale and then these penalties are summed.
	\item \textbf{Usage of the tonic.} A positive reward is assigned to a roll as a whole if share of columns where the lowest sounding note belongs to the tonic pitch class lies within a specified interval.  
	\item \textbf{Conjunct motion.} A positive reward is granted for every sounding cell such that the previous cell of the same pitch is not sounding, but there is a sounding cell of close enough pitch in the previous column.
	\item \textbf{Contrary motion.} A positive reward is granted for every column such that there are both a sounding cell of pitch lower than that of closest to it sounding cell from the previous column and a sounding cell of pitch higher than that of closest to it sounding cell from the previous column.
	\item \textbf{Lines.} Ideally, a created composition should be decomposable into melodic lines. To have $k$ melodic lines, it is necessary (but not sufficient) that exactly $k$ cells are sounding in every column. So below penalty may be used as a proxy for scoring presence of desired number of melodic lines:
	$$-\sum_{t = 1}^m \left| k - \sum_{i=1}^n R_{it} \right|.$$  
    \item \textbf{Noncyclicity.} Pieces where the same sequence of sound events is repeated over and over again should be penalized or, as a matter of choice, pieces should be rewarded for their noncyclicity. Here, the latter approach is used. The reward is implemented as minimum number of positive cells over differences between piano roll and its shifted version for all shift values from 1 up to a certain value (currently, it is equal to 8). To be more precise, this score is also clipped from above in order to prevent rewarding of placing extra sounding cells. In literature, scores based on autocorrelation are sometimes used as alternative way to push results away from degenerate states.
    \item \textbf{Absence of pitches played for a long time.} Partial case of a degenerate roll is a roll with some pitches that are constantly sounding. A negative penalty is assigned to every sounding cell such that cells of the same pitch are sounding in all $l$ previous columns where $l$ is a hyperparameter.
\end{itemize}

@@

To use less computation time, size of piano roll was chosen as follows: $n = 25$ (two octaves and one more note) and $m = 33$ (note that the last time step is silent by design, so only 32 time steps are filled). Training an agent from scratch on a CPU of a regular laptop takes less than an hour with these settings.

@@

If setup become more complicated after above modifications, it is worth trying to use LSTM neural network as actor model and evolutionary strategies as training algorithm.

>>>>

\date{November 2, 2019}

@@

\keywords{algorithmic composition \and music generation \and reinforcement learning}

@@

Currently, there are no perfect tools for algorithmic composition. Plenty of various approaches for composing music automatically exist, but none of them produce well-structured and novel pieces that cannot be distinguished from works of a talented human composer. Thus, there is an open research problem. Several recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating, payne2019musenet} and reinforcement learning \cite{smith2012reinforcement} are applicable. Moreover, some researchers combine them \cite{jaques2016generating, kotecha2018bach, kumar2019polyphonic}.

@@

Agent takes action by drawing it from probability distribution returned by its actor model. An actor model can be any function that takes observation as input and returns probability distribution over cells from the current time step. Here, shallow neural network is used as actor model. Its weights can be viewed as something that defines transition matrix for a Markov chain from which sequence of piano roll's states is generated.

@@

\begin{itemize}
	\item \textbf{Consonances.} Speaking formally, total reward for consonances is:
	$$\sum_{t = 1}^m \sum_{i, j = 1}^n [i > j]R_{it} R_{jt} c((i - j) \bmod 12),$$
	where square brackets are Iverson brackets and $c$ is a function that maps interval in semitones into measure of its consonance, $c: \{0, \dots, 11\} \to \mathbb{R}$.
	The interpretation is as follows. Since $R_{it} \in \{0, 1\}$, only pairs of sounding cells from the same column affect the total sum. Value of each non-zero term depends on interval between the two corresponding pitches. The more consonant the interval is, the higher the value should be. Function $c$ must be chosen in a way providing that. However, there is an exception -- here, $c(0) = 0$ although intervals of whole number of octaves are perfectly consonant. This is done in order to prevent doubling the same melody in different octaves. 
	\item \textbf{Absence of notes not from the specified scale.} Mode (major or minor) and position of a row with the tonic are set by a user before training. The same negative penalty is assigned to each sounding cell of a pitch that does not belong to the specified scale and then these penalties are summed.
	\item \textbf{Usage of the tonic.} A positive reward is assigned to a roll as a whole if share of columns where the lowest sounding note belongs to the tonic pitch class lies within a specified interval.  
	\item \textbf{Conjunct motion.} A positive reward is granted for every sounding cell such that the previous cell of the same pitch is not sounding, but there is a sounding cell of close enough pitch in the previous column.
	\item \textbf{Contrary motion.} A positive reward is granted for every column such that there are both a sounding cell of pitch lower than that of the closest to it sounding cell from the previous column and a sounding cell of pitch higher than that of the closest to it sounding cell from the previous column.
	\item \textbf{Lines.} Ideally, a created composition should be decomposable into melodic lines. To have $k$ melodic lines, it is necessary (but not sufficient) that exactly $k$ cells are sounding in every column. So below penalty may be used as a proxy for scoring presence of desired number of melodic lines:
	$$-\sum_{t = 1}^m \left| k - \sum_{i=1}^n R_{it} \right|.$$  
    \item \textbf{Noncyclicity.} Pieces where the same sequence of sound events is repeated over and over again should be penalized or, as a matter of choice, pieces should be rewarded for their noncyclicity. Here, the latter approach is used. The reward is implemented as minimum number of positive cells over differences between piano roll and its shifted version for all shift values from 1 up to a certain value (currently, it is equal to 8). To be more precise, this score is also clipped from above in order to prevent rewarding of placing extra sounding cells. In literature, scores based on autocorrelation are sometimes used as alternative way to push results away from degenerate states.
    \item \textbf{Absence of pitches played for a long time.} Partial case of a degenerate roll is a roll with some pitches that are constantly sounding. A negative penalty is assigned to every sounding cell such that cells of the same pitch are sounding in all $l$ previous columns where $l$ is a hyperparameter.
\end{itemize}

@@

To use less computation time, size of piano roll was chosen as follows: $n = 25$ (two octaves and one more note) and $m = 33$ (recall that the last time step is silent by design, so only 32 time steps are filled). Training an agent from scratch on a CPU of a regular laptop takes less than an hour with these settings.

@@

If setup becomes more complicated after above modifications, it might be worth trying to use LSTM neural network as actor model and evolutionary strategies as training algorithm.
